---
title: "R and Python in one post"
description: "Cleaning penguin data in R, summarising it in Python, and plotting the result back in R, all in one Quarto document."
date: 2026-09-23
categories: [R, Python, data]
engine: knitr
---

R and Python are often treated as rivals, but they can share one document.
Here I use each for the part it does comfortably: R to load and tidy the
Palmer Penguins data, Python (`pandas`) to summarise it, and R again to plot
the result. The [`reticulate`](https://rstudio.github.io/reticulate/) package
hands the objects back and forth.

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
Palmer Station Antarctica LTER (CC0 licence). Loaded from the `palmerpenguins`
R package.

## Pointing R at the Python environment

`reticulate` needs to know which Python to use. I point it at the `uv`
environment in `.venv`, so the post uses the same packages as the Python post.

```{r}
#| label: setup
library(reticulate)

venv_python <- file.path(Sys.getenv("QUARTO_PROJECT_DIR", "."), ".venv", "bin", "python")
if (file.exists(venv_python)) use_python(venv_python, required = TRUE)

py_config()$version
```

## R: load and tidy

```{r}
#| label: r-tidy
#| message: false
library(dplyr)
library(palmerpenguins)

penguins_clean <- penguins |>
  filter(!is.na(body_mass_g), !is.na(flipper_length_mm))

dim(penguins_clean)
```

## Python: summarise

In a Python chunk, the R object is available as `r.penguins_clean`, and
`reticulate` has already converted it to a `pandas` DataFrame.

```{python}
#| label: py-summary
import pandas as pd

df = r.penguins_clean
print(type(df))

summary = (
    df.groupby("species", observed=True)
    .agg(n=("body_mass_g", "size"),
         mean_mass_g=("body_mass_g", "mean"),
         mean_flipper_mm=("flipper_length_mm", "mean"))
    .round(1)
    .reset_index()
)
summary
```

## R: plot what Python made

Back in R, the Python object is `py$summary`, which arrives as an ordinary
data frame.

```{r}
#| label: fig-summary
#| fig-cap: "Mean body mass by species, computed in Python and plotted in R."
library(ggplot2)

ggplot(py$summary, aes(x = species, y = mean_mass_g, fill = species)) +
  geom_col(show.legend = FALSE) +
  labs(x = "Species", y = "Mean body mass (g)") +
  theme_minimal()
```

@fig-summary crosses the language boundary twice: the data went from R to
Python, and the summary came back to R for plotting. Gentoo are again the
heaviest species by a wide margin.

```{r}
#| label: warn-demo
#| warning: true
# A deliberate warning, left visible on purpose: as.integer() on a species name
as.integer(as.character(py$summary$species[1]))
```

That warning is expected. It is a reminder that values passed between the
languages keep their types, so a text column will not quietly become a number.
