---
title: "Penguin body mass in R"
description: "Do heavier penguins have longer flippers? A short look at the Palmer Penguins data in R."
date: 2026-09-23
categories: [R, data]
---

Which of the three Palmer penguin species is the heaviest, and does body mass
track flipper length? I use R to find out.

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
Palmer Station Antarctica LTER (CC0 licence). Loaded from the `palmerpenguins`
R package.

## Loading and tidying

```{r}
#| message: false
#| warning: false
library(dplyr)
library(ggplot2)
library(palmerpenguins)

penguins_clean <- penguins |>
  filter(!is.na(body_mass_g), !is.na(flipper_length_mm))

nrow(penguins)
nrow(penguins_clean)
```

Removing rows with missing mass or flipper length drops a couple of birds, and
leaves the rest of the data intact.

## Mass by species

```{r}
penguins_clean |>
  group_by(species) |>
  summarise(
    n = n(),
    mean_mass_g = round(mean(body_mass_g)),
    sd_mass_g = round(sd(body_mass_g))
  )
```

Gentoo penguins are clearly the heaviest, about 1.4 kg more than Adelie or
Chinstrap on average. Adelie and Chinstrap are close to each other.

```{r}
#| label: fig-penguin-mass
#| fig-cap: "Body mass of penguins by species."
ggplot(penguins_clean, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = "Species", y = "Body mass (g)") +
  theme_minimal()
```

@fig-penguin-mass shows the same story: the Gentoo box sits well above the
other two with barely any overlap.

## Mass versus flipper length

```{r}
cor(penguins_clean$body_mass_g, penguins_clean$flipper_length_mm)
```

The correlation is strongly positive, so heavier penguins do tend to have
longer flippers. This is a quick look and not a model, but it matches what the
box plot suggested: much of the difference is between species, and Gentoo is
simply a bigger bird.
